{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2016-10-25 sftrends_manu.Rnw
author updates
through study site description

% aliased citations
% \defcitealias{FLDEP12}{FLDEP 2012}

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{din}[DIN]{dissolved inorganic nitrogen}
\acrodef{jas}[JAS]{July-August-September}
\acrodef{sfe}[SFE]{San Francisco Estuary}
\acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}
\acrodef{wwtp}[WWTP]{wastewater treatment plant}

@@

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ \Sexpr{raw}
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

\subsection{Study location and data}

@@

Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.  Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen inputs are considerable, several

POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends

The delta region...

>>>>

% aliased citations
\defcitealias{SutulaIR}{Sutula et al. in review}

%acronyms
\acrodef{chla}[chl-\textit{a}]{chlorophyll \textit{a}}
\acrodef{din}[DIN]{dissolved inorganic nitrogen}
\acrodef{emp}[EMP]{Environmental Monitoring Program}
\acrodef{iep}[IEP]{Interagency Ecological Program}
\acrodef{jas}[JAS]{July-August-September}
\acrodef{sfe}[SFE]{San Francisco Estuary}
\acrodef{wrtds}[WRTDS]{Weighted Regressions on Time, Discharge, and Season}
\acrodef{wwtp}[WWTP]{wastewater treatment plant}

@@

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$, Phil Trowbridge$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ \Sexpr{raw}
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

\subsection{Study system}

@@

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.  Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \citep{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

\subsection{Data sources}

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with adequate QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on the continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

Daily flow estimates for the Delta region were obtained from the

### 2016-10-25 sftrends_manu.tex
through study site description

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Tue Oct 18 17:52:14 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  Recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass, increased occurrence of hypoxic conditions, and increased abundance of phytoplankton species associated with harmful algal blooms.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers to the lower Bay.  Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.  Sources are not limited to the watershed, such that local inputs from wastewater and internal loading from sediment processes also represent important contributions.  Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have complicated the interpretation of trends in nutrient concentrations. Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.

Two data sources that provide multi-decadal time series of nutrients and daily flow records were used to develop a quantitative description fo

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$, Phil Trowbridge$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Thu Oct 20 16:04:44 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass \citep{Cloern07}, increased occurrence of hypoxic conditions \citepalias{SutulaIR}, and increased abundance of phytoplankton species associated with harmful algal blooms \citep{Lehman05,Lehman10}.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers \citep{Jassby00,Jassby08,Novick15}.  Quantitative descriptions of nutrient dynamics in the Delta are challenging given the numerous sources of nutrients and the volume of water that is exchanged through natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is lost.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \citep{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

\subsection{Data sources}

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with adequate QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on the continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

Daily flow estimates for the Delta region were obtained from the
